---
title: Lab 5
authors: Anthony Song, Jaydon Cobb
Github: aasong113, JaydonLCobb
---

```{r}
library(stringr)
library(readxl)
library(ggplot2)
library(maps)
library(ggmap)
library(dplyr)
library(magrittr)
library(tidyr)
library(tidyverse)
library("readxl")
```

```{r}
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
per <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```


Part one: Accident data

1.) are there some days of the week where more accidents happen than on others (use variable DAY_WEEK)?

###The median, mean and 3rd quartile values of the frequencies in which accidents occur on a particular day of the week are 4261, 4892 and 5359 accidents per day of the week respectively. Looking at the data table, we make the conclusions that days of the week that have frequencies larger than the 3rd quartile are days of the week that accidents occur more often than others. These days that have accident frequencies larger than the 3rd quartile are Saturay and Sunday. This makes sense because these are days of the week are where casual things occur, people are less focused, and this ultimately can cause accidents in operation of vehicles. 
```{r}
days_freq <- acc
days_freq <- data.frame(table(days_freq$DAY_WEEK))
days_freq$Var1 <- recode(days_freq$Var1, 
       "1"="Sunday",
       "2"="Monday",
       "3"="Tuesday",
       "4"="Wednesday",
       "5"="Thursday",
       "6"="Friday",
       "7"="Saturday")

summary(days_freq$Freq)
days_freq %>%
  ggplot(aes(x = Var1, y = Freq, fill = Var1)) + geom_bar(stat="identity")+labs(x = "Day of the Week", y = "Frequency", title = "Frequency of Accidents for a Particular Day of the Week", fill = "Day of the Week")
```

2.) what time of the day do accidents happen (use variable HOUR)?

### Similar to question one, I followed the same method to determine which hour of the day did accidents most frequently occur. According to the 5 number summary of the frequencies, we can see that the median, mean and 3rd quartile are 1321, 1370 and 1760 respectively. Looking at our barchart and frequency table we can see that the hours: 16, 17, 18, 19, 20 and 21 had frequencies larger than the 3rd quartile. This makes sense because at these hours people are off work performing leisure activities. Their focus is less guarded and it is darker outside, this combination can lead to car accidents. 
```{r}
hour_freq <- acc
hour_freq <- data.frame(table(hour_freq$HOUR))
hour_freq$Var1 <- recode(hour_freq$Var1, "99"="N/A")
summary(hour_freq$Freq)
hour_freq %>%
  ggplot(aes(x = Var1, y = Freq, fill = Var1)) + geom_bar(stat="identity")+labs(x = "Hour of the Day", y = "Frequency", title = "Frequency of Accidents for a Particular Hour of the Day", fill = "Hour of the Week") 

```

3.) what is the number of accidents with at least one drunk driver (use variable DRUNK_DR)?

###The number of accidents with at least one drunk driver is 8769 accidents. 
```{r}
dr_drive <- acc
dr_drive <- data.frame(table(dr_drive$DRUNK_DR))
num_dr_drive <- sum(dr_drive[2:4, 2])
num_dr_drive
```

4. In the person table, identify drivers (PER_TYP == 1, see FARS manual) and subset on them.

### To identify the drivers we select the rows which PER_TYP == 1. This subsets our dataset from 84921 rows to 52274 rows.
```{r}
drivers_only <- per
drivers_only <- drivers_only[which(drivers_only$PER_TYP == 1),]
drivers_only <- na.omit(drivers_only)
```

5. Join accident and person table (work out which variable(s) to use)

Variables to use (temp_acc):persons, city,Month,day,  hour, TWAY_ID, LATITUDE, LONGGITUD, WEATHER,FATALS

Variables to use (temp_drive):Month,day,  hour,sex, per_typ, age, drugs, County
```{r}
#nrow(acc) = 34247
temp_drive <- drivers_only %>%
  distinct(ST_CASE, .keep_all = TRUE) 
#nrow(temp_drive) = 34157

# Find the differences in the two data sets. 
differences <- setdiff(acc$ST_CASE, temp_drive$ST_CASE)
# length(differences) = 90 

# remove values that are in the differences vector. 
temp_acc <- acc
temp_acc <- temp_acc[ ! temp_acc$ST_CASE %in% differences, ]
# nrow(temp_acc) == nrow(temp_drive) is true. 

# Extract certain columns
acc_col_extract <- temp_acc %>%
  select(PERSONS, CITY, MONTH, DAY, HOUR, TWAY_ID, LATITUDE, LONGITUD, WEATHER, FATALS, DAY_WEEK)
per_col_extract <- temp_drive %>%
  select(SEX, PER_TYP, AGE, DRUGS, COUNTY)
# merge the data frames. 
joint_dat <- cbind(acc_col_extract, per_col_extract)
```


6. Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results.
```{r}

joint_dat$DAY_WEEK <- recode(joint_dat$DAY_WEEK, 
       "1"="Sunday",
       "2"="Monday",
       "3"="Tuesday",
       "4"="Wednesday",
       "5"="Thursday",
       "6"="Friday",
       "7"="Saturday")

joint_dat$SEX <- recode(joint_dat$SEX, 
       "1"= "Male", "2" = "Female", "8" = "Not Reported", "9" = "Unknown")

joint_dat_day <- data.frame(table(joint_dat$DAY_WEEK))


joint_dat_hour <- data.frame(table(joint_dat$HOUR))


joint_dat_sex <- data.frame(table(joint_dat$SEX))

joint_box <- joint_dat
joint_box <- joint_box[!joint_box$HOUR==99,]
joint_box <- joint_box[!joint_box$SEX=="Not Reported",]
joint_box <- joint_box[!joint_box$SEX=="Unknown",]

### visualizing all of the data. 
ggplot(joint_box) + geom_boxplot( aes(x = DAY_WEEK, y = HOUR )) + facet_wrap(~SEX)+ theme(axis.text.x = element_text(angle = 60, hjust = 1))+labs(x = "Day of the Week", y = "Hour of the Day", title = "Boxplot of the Hour of the Day, for Each Day of the Week, by Reported Sex")

### Bar charts to tally the number visually. 
joint_dat_day %>%
  ggplot(aes(x = Var1, y = Freq, fill = Var1)) + geom_bar(stat="identity")+labs(x = "Day of the Week", y = "Frequency", title = "Frequency of Accidents for a Particular Day of the Week", fill = "Day of the Week")

joint_dat_hour %>%
  ggplot(aes(x = Var1, y = Freq, fill = Var1)) + geom_bar(stat="identity")+labs(x = "Hour of the Day", y = "Frequency", title = "Frequency of Accidents for a Particular Hour of the Day", fill = "Hour of the Week") 

joint_dat_sex %>%
  ggplot(aes(x = Var1, y = Freq, fill = Var1)) + geom_bar(stat="identity")+labs(x = "Sex", y = "Frequency", title = "Frequency of Accidents for a Particular Sex", fill = "Sex") 

```
###use readxl::read_xlsx to read in the GLCs
```{r}
glc <- readxl::read_xlsx("FRPP GLC - United StatesFeb132020.xlsx")
```
### Tidy up the glc data. 
```{r}
### rename
glc<- glc %>%
  rename(
    "STATE" = "State Code",
    "COUNTY" = "County Code"
  )

drops <- c("Country Code","Old City Name", "Date Record Added")
glc <- glc[ , !(names(glc) %in% drops)]


glc <- transform(glc, STATE = as.numeric(STATE), 
               COUNTY = as.numeric(COUNTY))
glc
```


```{r}

df <- glc %>%
  select('State.Name', STATE = 'STATE', COUNTY = 'County.Name', 'County.Name') %>%
  unique()
df$STATE <- as.integer(df$STATE)
df$COUNTY <- as.integer(df$STATE)


df1 <- df %>%
  select(STATE, COUNTY) %>%
  group_by(STATE, COUNTY) %>%
  summarize(total_accidents= n()) %>%
  left_join(glc, by = c('STATE', 'COUNTY'))

df1

df2 <- map_data('county')

df2 <- df2 %>%
  mutate(`State Name` = region, `County Name` = subregion) %>%
  select(`State Name`, `County Name`, long, lat, group, order)

df1 <- df1 %>%
  full_join(df1, by = c('State.Name', 'County.Name'))

df1
```




7/8. Now plot a choropleth map of accidents on a county level. Visualize the number of accidents in each county using a choropleth map. To do this, you need to join the accident data, GLCs, and appropriate map data.

###The county ID is accounted for by color. The abundance of dots, with their color corresponding to country ID, represents the number of accidents in a certain region.
```{r}
# merge two data frames by STATE and COUNTY
temp_acc <- acc

merged <- merge(data.frame(temp_acc, row.names=NULL), data.frame(glc, row.names=NULL), by = 0, all = TRUE)[-1]
merged_complete <- na.omit(merged)

# graph
graph_dat <- merged_complete

# Get outline of US maps
states <- map_data("state")

# obtain only mainland state values
graph_dat <- graph_dat[!graph_dat$COUNTY.x==0,]
graph_dat <- graph_dat[!graph_dat$COUNTY.x==999,]
graph_dat <- graph_dat[!graph_dat$COUNTY.x==997,]
graph_dat <- graph_dat[!graph_dat$COUNTY.x==998,]
graph_dat<-graph_dat[graph_dat$LONGITUD < 0,]
graph_dat<-graph_dat[graph_dat$LONGITUD > -130,]
graph_dat<-graph_dat[graph_dat$LATITUDE < 50,]


states %>% ggplot(aes(x = long, y = lat)) + 
  geom_path(aes(group = group))+geom_point(data = graph_dat, aes(x = LONGITUD, y = LATITUDE, color = COUNTY.x), alpha = 0.7, size = 0.02) +
  labs(x = "Longitude", y = "Latitude", title = "Choropleth Map of the US: Accidents by County")

```



9. Can you find seasonal trends in accident occurrences in different states? Is summer or winter more dangerous? Why?

### Using the choropleth map we are able to visualize some seasonal trends in which accidents occur. As expected the winter months have the largest amount of accidents, especially along the coastlines of the northern lake states like Wisconsin and Michigan. This can be seen by the abundance of yellow points. But it is important to notice, that in the months building up to the winter (September and October) there are also a lot of points, as shown by light green. This shows that the build up to the winter produces a lot of accidents as well. But not to be overshadowed the summer months also show a large showing accident points in the congested areas like the east coast. This may be due to the driving culture and congested highways and speedways.

### In hindsight we may think that the winter is far more dangerous than the summer due to slippery roads and harsh conditions. But when looking at the frequency of accidents that occur during each season we can see that their difference is not extraordinary. Although the winter is considered more dangerous due to a higher frequency of accidents, this is most likely caused by the harsh driving conditions. It should be noted that the summer is the second ahead of the spring and fall. This may be due to excessively fast drivers in the summer and more carelessness, causing less focus on the road. 

```{r}
#Encode months
graph_dat$MONTH <- recode(graph_dat$MONTH, 
       "1"="January",
       "2"="February",
       "3"="March",
       "4"="April",
       "5"="May",
       "6"="June",
       "7"="July",
       "8" = "August",
       "9" = "September",
       "10"="October",
       "11"="November",
       "12" = "December")
# as factor
graph_dat$MONTH = factor(graph_dat$MONTH,levels=c("January","February","March",
               "April","May","June","July","August","September",
               "October","November","December"),ordered=TRUE)

states %>% ggplot(aes(x = long, y = lat)) + 
  geom_path(aes(group = group))+geom_point(data = graph_dat, aes(x = LONGITUD, y = LATITUDE, color = MONTH), alpha = 0.7, size = 0.1) +
  labs(x = "Longitude", y = "Latitude", title = "Choropleth Map of the US: Accidents by Month of the Year")
```

```{r}
#Encode seasons 
graph_dat2 <- graph_dat
graph_dat2$MONTH <- recode(graph_dat$MONTH, 
       "January" = "Winter",
       "February" = "Winter",
       "March" = "Spring",
       "April" = "Spring",
       "May" = "Spring",
       "June" = "Summer",
       "July" = "Summer",
       "August" = "Summer",
       "September" = "Fall",
       "October" = "Fall",
       "November" = "Winter",
       "December" = "Winter")
seasons <- data.frame(table(graph_dat2$MONTH))

seasons %>%
  ggplot(aes(x = Var1, y = Freq , fill = Var1)) + geom_bar(stat = "identity")+labs(x = "Seasons", y = "Frequency", title = "Frequency of Accidents for a Particular Season", fill = "Season") 


```

# The number of accidents balances over the seasons, we do have a higher number in the winter which is to be expected but overall across all 4 seasons we can see the amount of accidents balance out.
